\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add final clarifications

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to sample from the study base a
finite amount of \emph{person moments}. These person moments are indexed
by both an individual in the study and a time point, and therefore each
person moment has a covariate profile, an exposure status and an outcome
status attached to it. We note that there is only a finite number of
person moments associated with the event of interest (what Hanley \&
Miettinen call the \emph{case series}). Case-base sampling refers to the
sampling from the base of a representative finite sample called the
\emph{base series}.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]

Next, we model the base series sampling mechanism using non-homogeneous
Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\), with the
person-moments where \(dR_i(t) = 1\) constituting the base series. The
process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
base series person-moments contributed by individual \(i\). This process
is typically defined by the user via its intensity function
\(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).

@@

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package
\CRANpkg{cmprsk} provides methods for cause-specific subdistribution
hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF}
below). Packages providing this functionality include \pkg{CFC},
\pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{riskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the survival function, whereas the second category
often produces estimates in the form of stepwise functions.

@@

\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline

@@

For the third case study, we show how \pkg{casebase} can also be used
for variable selection through regularized estimation of the hazard
function as given by \eqref{eq:penest}. To illustrate this
functionality, we use the dataset from the Study to Understand Prognoses
Preferences Outcomes and Risks of Treatment (SUPPORT)
\citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
The SUPPORT dataset tracks death in five American hospitals within
individuals who are considered seriously ill.~The cleaned and imputed
data consists of 9104 observations and 30 variables, and it is available
as part of the \pkg{casebase} package. In the comparisons below, all
covariates except \code{sps} and \code{aps} were used. These two
variables correspond to scores for predicting the outcome that were
developed as part of the original study. For more information about this
dataset, the reader is encouraged to look at the documentation in our
package.

For our penalized case-base model, we opt for the natural log of time
which corresponds to a Weibull distribution. For fitting the penalized
hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
to the \code{fitSmoothHazard} function. We supply both a matrix \code{y}
containing the time and event variables, and a matrix \code{x}
containing all other covariates. We apply the lasso penalty by setting
\code{alpha = 1} and assign a \code{penalty.factor} (\(w_j\);
cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is
in the selected model. We compare our approach to both Cox regression,
and lasso penalized Cox regression (fitted via the \pkg{glmnet}
package).

>>>>

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person moments}. These person moments
are indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure
status, and an outcome status attached to it. By comparing person
moments at which an event occurred with person moments at which no event
occurred, we can extract information about hazards and survival.

Therefore, we start by sampling all person moments at which the event
occurred; this collection of person moments is what Hanley \& Miettinen
call the \emph{case series}. The incidence of the case series is
dictated by the hazard function of interest. Next, we sample a finite
number of person moments at which no event occurred; this second
collection of person moment is what Hanley \& Miettinen call the
\emph{base series}. The sampling mechanism for the base series is left
at the discretion of the user, but in practice we find that sampling
uniformly from the study base provides both simplicity and good
performance. This is the default sampling mechanism in the package.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\] The processes
\(N_i(t)\) count the person moments from the \emph{case series}.

To complement the case series, we sample person moments for the base
series. To do so, we model the base series sampling mechanism using
non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
the person-moments where \(dR_i(t) = 1\) constitute the base series. The
process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
base series person-moments contributed by individual \(i\). The
processes \(R_i(t)\) are typically defined by the user via its intensity
function \(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).

@@

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
provides functionality for competing-risk analysis and multistate
modelling. The package \CRANpkg{cmprsk} provides methods for
cause-specific subdistribution hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIFs.

Finally, several packages include functions to estimate the survival
function and the CIF. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explicitly modeled. Using this estimate, the survival function and
the CIF can be obtained using their functional relationships (see
Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and
\pkg{survival}. Our package \pkg{casebase} also follows this approach
for both single-event and competing-risk analyses. The second category
outlined above broadly corresponds to semi-parametric models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator) or parametric estimators (e.g.~spline
functions). Packages that implement this approach include
\pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the survival function, whereas the second category often produces
estimates in the form of stepwise functions.

@@

\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &  \checkmark              \\ \hline

@@

For the third case study, we show how \pkg{casebase} can also be used
for variable selection through regularized estimation of the hazard
function as given by \eqref{eq:penest}. We note that this is different
than the semiparametric model Coxnet, which regularizes the Cox partial
likelihood. To illustrate this functionality, we use the dataset from
the Study to Understand Prognoses Preferences Outcomes and Risks of
Treatment (SUPPORT)
\citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
The SUPPORT dataset tracks death in five American hospitals within
individuals who are considered seriously ill.~The cleaned and imputed
data consists of 9104 observations and 30 variables, and it is available
as part of the \pkg{casebase} package. In the comparisons below, all
covariates except \code{sps} and \code{aps} were used. These two
variables correspond to scores for predicting the outcome that were
developed as part of the original study. For more information about this
dataset, the reader is encouraged to look at the documentation in our
package.

For our penalized case-base model, we opt for the natural log of time
which corresponds to a Weibull distribution. For fitting the penalized
hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
to the \code{fitSmoothHazard} function. The \code{fitSmoothHazard} and
\code{fitSmoothHazard.fit} functions sample the case and base series,
calculate the required offset, and transform the data to match the
expected input of the \pkg{glmnet} package. The penalized logistic
regression is then fit for multiple values of the tuning parameter using
the function \code{glmnet::cv.glmnet} and the binomial family. We supply
both a matrix \code{y} containing the time and event variables, and a
matrix \code{x} containing all other covariates. We apply the lasso
penalty by setting \code{alpha = 1} and assign a \code{penalty.factor}
(\(w_j\); cf.~Equation \ref{eq:penest}) of 0 to the time variable to
ensure it is in the selected model. We compare our approach to both Cox
regression, and lasso penalized Cox regression (fitted via the
\pkg{glmnet} package and using the Cox family).
